Learning and comparing functional connectomes across subjects

### 2013-01-04 paper.tex

\paragraph{FMRI-based function definition}
%
Defining regions directly from the fMRI signal brings many benefits.
First, it can capture subject-specific functional information. Second, it
adapts to the signal at hand and its limitations, such as
image distortions or vascular and movement artifacts that are isolated
in ICA-like approaches. Lastly, incorporating functional information into
regional definition will result in more homogenous regions that better represent
connectivity present at the voxel level then anatomically defined atlases such
as AAL or Harvard-Oxford \cite{craddock2012}. The simplest approach to define task-specific
regions is to use activation maps derived from standard GLM-based
analysis in a task-driven study (see for instance \cite{poldrack2011}).
Regions are extracted by thresholding the maps, or using balls around the
activation peaks. For resting-state studies, unsupervised multivariate
analysis techniques are necessary. Clustering approaches extract
full-brain parcellations \cite{craddock2012, bellec2010, yeo2011,
thirion2006}, and have been shown to segment well-known functional
structures from rest data. Alternatively, decompositions
methods, such as ICA \cite{beckmann2004}, can unmix linear
combinations of multiple effects and separate out overlapping spatial maps
capturing functional networks or confounding effects. At high model
order, ICA maps define a functional parcellation \cite{kiviniemi2009}.
Extracting regions from these maps requires additional effort as they can
display fragmented spatial features and structured background noise, but
incorporating sparsity and spatial constraints in the decomposition
techniques leads to contrasted maps that outline many different
structures \cite{craddock2012, varoquaux2012} (see fig.\,\ref{fig:parcellations}).

@@

\paragraph{Signal extraction}
%
Given a set of graph nodes, the next step is to extract a representative time
series for each node.  To study \emph{intrinsic} activity, \emph{e.g.}\ with
rest data, signal extraction can be achieved by either averaging the fMRI time
series across the voxels in a region, or by taking the first eigenvariate from a
principle components analysis of the time series \cite{friston2006}.
Comparisons of these methods has shown that the eigenvariate method is
more sensitive to function inhomogeneity \cite{craddock2012} and exhibits worse
test-retest reliability than averaging time series \cite{zuo2010}. 
In addition, improved specificity to BOLD signal can be enforced
by using only signal in voxels near gray-matter tissues. For this purpose, we
suggest summarizing the signal in an ROI by a mean of the different voxels
weighted by the subject-specific gray matter probabilistic segmentation, as
output by \emph{e.g.}\ SPM's segmentation tool \cite{ashburner2005}.

@@

\paragraph{Graph-topology metrics}
%
Functional connectivity graphs have been found to display specific
topological properties\footnote{In the neuroscience world, these
descriptions are grouped under the terms of ``graph theoretical
approaches'', however graph theory is an entire division of mathematics
and computer science that is concerned with much more than topology of
random graphs.} that are characteristic of small-word networks
\cite{stam2004,salvador2005,achard2006,bullmore2009}. These networks
display excellent transport properties: although they have a relatively
small number of connections, any two regions of the brain are well
connected. 
Another interesting consequence of their specific topology is
the resilience it gives the system to attacks such as resulting from
brain lesions \cite{achard2006}. This overall structure of
functional-connectivity graphs can be summarized by a few metrics, such
as the average path length between any two nodes or the local clustering
coefficients \cite{rubinov2010}. Given that pathologies without a
localized focus, such
as schizophrenia, are thought to have a global impact on brain
connectivity \cite{liu2008,bassett2008}, the graph-topology metrics are
promising markers to perform inter-subject comparison. Such an
approach is appealing as it is not subject to multiple comparison issues.
However, it has been criticized as giving a fairly unspecific
characterization of the brain and being fragile to noise
\cite{ioannides2007}. Another caveat is that these properties are not
specific to brain function: correlation matrices display
small-word properties such as local clustering by construction. Indeed, if two
nodes are strongly correlated to a third, they are highly likely to be
correlated to each other \cite{zalesky2012}. This observation highlights
the need for well defined null-hypothesis \cite{zalesky2012,rubinov2011},
but also for controlled recovery of brain functional connectivity going
beyond empirical correlation matrices, as discuss in the previous section.
% CAMERON What about regional summary measures such as centrality??? Xinian's work
% maybe a paragraph that covers ReHo, degree centrality, eigenvector centrality, pagerank etc...

@@

Predictive modeling is concerned with learning (or fitting) a model that is
capable of decoding or \emph{predicting} information from new unseen data
\cite{pereira2009}.  In the context of connectomes, predictive modeling provides
a means for delineating connectivity based biomarkers of disease diagnosis,
prognosis, or other phenotypic outcomes \cite{craddock2009, dosenbach2010}. The
accuracy of a predictive model provides a measure of the amount of information
present in the connectome about the phenotypic measure being evaluated
\cite{kjems2002, kriegeskorte2006}.  When combined with reproducibility, prediction accuracy
provides a metric for evaluating experimental trade-offs for data acquisition,
preprocessing, and analysis \cite{strother2002, laconte2003}.  Multivariate
predictive modeling (MPM)
is attractive in connectomics because it is sensitive to dependencies
between features and avoids the need to correct for multiple comparisons since
the significance of an entire pattern is evaluated using a single statistical
test. 
Additionally many MPM techniques exist which are specifically
formulated to deal with high dimensional datasets with few observations.
Predictive modeling techniques have been
successfully applied to identify connectome-based biomarkers of Alzheimer's
disease \cite{stonnington2010}, depression \cite{craddock2009, zeng2012}, schizophrenia
\cite{cecchi2009, shen2010}, autism \cite{anderson2011}, ADHD \cite{zhu2008}, aging
\cite{dosenbach2010}, as well as to classify mental operations
\cite{richiardi2011,shirer2012}. The growing interest for applying MPM 
to connectivity analysis was highlighted by the ADHD200
Global Competition, in which the object was to identify a connectivity based
biomarker of ADHD \cite{adhd2002012}. Recent work has illustrated the utility
of predictive modeling for deriving connectivity models at the individual level
\cite{chu2011}.

@@

Although MPM techniques are extremely well suited at measuring whether
information exists in the connectome about a phenotypic variable, identifying
the connections most relevant to the predictive model (feature selection) is
less straight forward.  These problems can be somewhat mitigated by carefully
previously described sparse inverse covariance estimation techniques to minimize
the number of connections. Feature selection \cite{guyon2003} can be performed by filtering out
features based on their statistical relationship with the variable of interest
\cite{craddock2009, shen2010},
selecting features based on cross-validation \cite{craddock2009, deshpande2010}, or by employing sparse modeling
methods that force weights for non-essential features to be zero
\cite{tibshirani1994, michel2009, ryali2010}.
Filtering features based on a univariate criterion such as t-test should be
avoided, since they are insensitive to multivariate information and may exclude
features that could be important to a multivariate model \cite{guyon2003,
	craddock2009}. No matter which
feature selection strategy is employed, it must be performed within
cross-validation to avoid biasing estimates of prediction accuracy.

The interpretation of connections implicated by MPM and their relationship with
a phenotypic outcome is complicated and requires insight into the mechanisms
underlying the modeling approach. If features are appropriately standardized
prior to training, the magnitude of model weights can be interpreted to reflect
the relative importance of the corresponding feature to the model, similar to
beta values from a standard regression model. But due to the multivariate nature
of the MPMs, the sign of the weight cannot necessarily be interpreted to
indicate the relationship between a connection and a phenotypic variable.
Indeed, it is possible for the sign of a model weight to change based on the
inclusion or exclusion of another feature in the model \cite{craddock2009}.  It
is perhaps most reasonable to adapt a conservative interpretation in which
predictive modeling is used to identify candidate connections that are later
tested in follow-on experiments better suited to elucidating their relationship
to the variable of interest.

>>>>

\paragraph{FMRI-based function definition}
%
Defining regions directly from the fMRI signal brings many benefits.
First, it can capture subject-specific functional information. Second, it
adapts to the signal at hand and its limitations, such as
image distortions or vascular and movement artifacts that are isolated
in ICA-like approaches. Lastly, incorporating functional information into
regional definition will result in more homogenous regions that better represent
connectivity present at the voxel level then anatomically defined atlases such
as AAL or Harvard-Oxford \cite{craddock2012}. The simplest approach to define task-specific
regions is to use activation maps derived from standard GLM-based
analysis in a task-driven study (see for instance \cite{poldrack2011}).
Regions are extracted by thresholding the maps, or using balls around the
activation peaks. For resting-state studies, unsupervised multivariate
analysis techniques are necessary. Clustering approaches extract
full-brain parcellations \cite{craddock2012,bellec2010,yeo2011,thirion2006}, and have been shown to segment well-known functional
structures from rest data. Alternatively, decompositions
methods, such as ICA \cite{beckmann2004}, can unmix linear
combinations of multiple effects and separate out overlapping spatial maps
capturing functional networks or confounding effects. At high model
order, ICA maps define a functional parcellation \cite{kiviniemi2009}.
Extracting regions from these maps requires additional effort as they can
display fragmented spatial features and structured background noise, but
incorporating sparsity and spatial constraints in the decomposition
techniques leads to contrasted maps that outline many different
structures \cite{varoquaux2012} (see fig.\,\ref{fig:parcellations}).
structures \cite{varoquaux2012} (see fig.\,\ref{fig:parcellations}).

@@

\paragraph{Signal extraction}
%
Given a set of graph nodes, the next step is to extract a representative time
series for each node.  To study \emph{intrinsic} activity, \emph{e.g.}\ with
rest data, signal extraction can be achieved by either averaging the fMRI time
series across the voxels in a region, or by taking the first eigenvariate from a
principle components analysis of the time series \cite{friston2006}.
Comparisons of these methods has shown that the eigenvariate method is
more sensitive to function inhomogeneity \cite{craddock2012} and exhibits worse
test-retest reliability than averaging time series \cite{zuo2010}. 
In addition, improved specificity to BOLD signal can be enforced
by using only signal in voxels near gray-matter tissues. For this purpose, we
suggest summarizing the signal in an ROI by a mean of the different voxels
weighted by the subject-specific gray matter probabilistic segmentation, 
as output by \emph{e.g.}\ SPM's segmentation tool \cite{ashburner2005} or
FSL's FAST program \cite{zhang2001}.

@@

\paragraph{Graph-topology metrics}
%
Functional connectivity graphs have been found to display specific
topological properties\footnote{In the neuroscience world, these
descriptions are grouped under the terms of ``graph theoretical
approaches'', however graph theory is an entire division of mathematics
and computer science that is concerned with much more than topology of
random graphs.} that are characteristic of small-word networks
\cite{stam2004,salvador2005,achard2006,bullmore2009}. These networks
display excellent transport properties: although they have a relatively
small number of connections, any two regions of the brain are well
connected. 
Another interesting consequence of their specific topology is
the resilience it gives the system to attacks such as resulting from
brain lesions \cite{achard2006}. This overall structure of
functional-connectivity graphs can be summarized by a few metrics, such
as the average path length between any two nodes, the local clustering
coefficients, or the node degree centrality \cite{rubinov2010}. Given that pathologies without a
localized focus, such
as schizophrenia, are thought to have a global impact on brain
connectivity \cite{liu2008,bassett2008}, the graph-topology metrics are
promising markers to perform inter-subject comparison. Such an
approach is appealing as it is not subject to multiple comparison issues.
However, it has been criticized as giving a fairly unspecific
characterization of the brain and being fragile to noise
\cite{ioannides2007}. Another caveat is that these properties are not
specific to brain function: correlation matrices display
small-word properties such as local clustering by construction. Indeed, if two
nodes are strongly correlated to a third, they are highly likely to be
correlated to each other \cite{zalesky2012}. This observation highlights
the need for well defined null-hypothesis \cite{zalesky2012,rubinov2011},
but also for controlled recovery of brain functional connectivity going
beyond empirical correlation matrices, as discuss in the previous section.
% CAMERON What about regional summary measures such as centrality??? Xinian's work
% maybe a paragraph that covers ReHo, degree centrality, eigenvector centrality, pagerank etc...

@@

Predictive modeling is concerned with learning (or fitting) a model that is
capable of \emph{predicting} information from new unseen data
\cite{pereira2009}.  In the context of connectomes, predictive modeling provides
a means for delineating connectivity based biomarkers of disease diagnosis,
prognosis, or other phenotypic outcomes \cite{craddock2009, dosenbach2010}. The
accuracy of a predictive model provides a measure of the amount of information
present in the connectome about the phenotypic measure being evaluated
\cite{kjems2002, kriegeskorte2006}.  When combined with reproducibility, prediction accuracy
provides a metric for evaluating experimental trade-offs for data acquisition,
preprocessing, and analysis \cite{strother2002, laconte2003}.  Multivariate
predictive models are attractive in connectomics because they are sensitive to dependencies
between features and avoid the need to correct for multiple comparisons since
the significance of an entire pattern is evaluated using a single statistical
test. 
Additionally modern predictive modeling techniques draw from the
statistical learning literature, which specifically addresses
 high dimensional datasets with few observations.
Predictive modeling has been
successfully applied to identify connectome-based biomarkers of Alzheimer's
disease \cite{stonnington2010}, depression \cite{craddock2009, zeng2012}, schizophrenia
\cite{cecchi2009, shen2010}, autism \cite{anderson2011}, ADHD \cite{zhu2008}, aging
\cite{dosenbach2010}, as well as to classify mental operations
\cite{richiardi2011,shirer2012}. The growing interest for applying
predictive modeling to connectivity analysis was highlighted by the ADHD200
Global Competition, in which the object was to identify a connectivity based
biomarker of ADHD \cite{adhd2002012}. Recent work has illustrated the utility
of predictive modeling for deriving connectivity models at the individual level
\cite{chu2011}.

@@

Although predictive modeling techniques are well suited at measuring whether
information exists in the connectome about a phenotypic variable, they do
not directly identify the connections that are most relevant to the prediction.  These problems can be somewhat mitigated  by
relying on previously described sparse inverse covariance estimation techniques to minimize
the number of connections. Feature selection \cite{guyon2003} 
can be performed by filtering out
features based on their statistical relationship with the variable of interest
\cite{craddock2009, shen2010}. Importantly, it must be performed within
cross-validation to avoid biasing estimates of prediction accuracy.
% 
The interpretation of connections used in predictive models and their
relationship with
a phenotypic outcome is difficult and requires insight into the mechanisms
underlying the modeling approach. For linear models, the weights of the
model are similar to the weights of
a ordinary linear regression. If features are appropriately standardized
prior to training, the magnitude of weights can be interpreted to reflect
the relative importance of the feature to the model, for instance the
corresponding edge in the connectome. However,
interpreting how the connections differs between classes or relates to a
phenotypic variable can be more complicated given the multivariate nature
of their involvement.
Indeed, it is possible for the sign of a model weight to change based on the
inclusion or exclusion of another feature in the model \cite{craddock2009}.  It
is perhaps most reasonable to adapt a conservative interpretation in which
predictive modeling is used to identify candidate connections that are later
tested in follow-on experiments better suited to elucidating their relationship
to the variable of interest.
